Algebraic Tori in Unitary Groups

### 2014-07-18 Paper.tex
Add more sketches of proofs

Fix a free $K$-module $V$ of dimension $2n$. There is a unique a non-degenerate alternating $K$-valued $K$-bilinear pairing $S$ on $V$.

$\Sp_S = \Sp = \Aut(V,S)$.

The forms of $(V,S)$ are in bijection with $H^1(\GalKbK, \Aut(V,S))$, as there is only a unique form, this cohomology group vanishes.

@@

For the purpose of this paper a symplectic group means a simply connected group of type $C_n$.
The outer automorphism group of such a group is trivial, the center if $\mu_2$.

We may construct the symplectic groups as defined above as follows. Fix a quaternion algebra $A$ over $K$ (we allow $M_2(K)$) together with the standard quaternionic involution $\sigma$. If $A$ is $M_2(K)$ then $\sigma( \left( \begin{smallmatrix} w & x\\ y & z \end{smallmatrix} \right) ) =\left( \begin{smallmatrix}z & {-x}\\ {-y} & w \end{smallmatrix} \right) $

@@

@@

\TODO{prove that this construction is universal - sketch follows}
\begin{itemize}
\item It suffices to prove that rational conjugacy is the same as rational isomorphism.
\item First claim any conjugate embedding (of groups) arises from a conjugate embedding (of algebras with involutions).
\item We want to show that any two embeddings $A \hookrightarrow C$ are conjugate.
        With $A,C$ as defined above but the embedding not necissarily coming from the condition.

@@

\TODO{Want to prove the following facts [that one hopes are facts]}
\begin{itemize}
\item The normalizer in $\Sp$ (resp. $\SO$) of $\U$ is $\mu_2 \rtimes \U$ (resp. $\U$).  [or something like this]
\subitem  suffices to check over the algebraic closure.
\item The centralizer of the center of $\U$ in $\Sp$ and $\SO$ is $\U$
\subitem suffices to check over the algebraic closure.
\subitem Note that the connected component of the normalizer of the center of $\U$ is the centralizer of the center.
              (first claim reduces to second + claim that normalizer is connected in $\SO$ case)
\end{itemize}

\TODO{Invariants of orthogonal from those of unitary}
\begin{itemize}
\item The discriminant of the unitary involution should determine clifford of the orthogonal.
\item The discriminant of the orthogonal should be determined by the choice of $E$.
\item The index of the unitary and orthogonal should be the same.
\end{itemize}

\TODO{Summary - Given orthogonal which unitary are possible}
\TODO{Summary - Given unitary which orthogonal are possible}

@@

>>>>

Fix a free $K$-module $V$ of dimension $2n$. There is a unique a non-degenerate alternating $K$-valued $K$-bilinear pairing $S$ on $V$.
We define the associated symplectic group to be $\Sp_S = \Sp = \Aut(V,S)$.
The forms of $(V,S)$ are in bijection with $H^1(\GalKbK, \Aut(V,S))$, as there is only a unique form, this cohomology group vanishes.

@@

For the purpose of this paper a symplectic group means a simply connected group of type $C_n$.
The outer automorphism group of such a group is trivial, the center is $\mu_2$.

We may construct these symplectic groups as follows. Fix a quaternion algebra $A$ over $K$ (we allow $M_2(K)$) together with the standard quaternionic involution $\sigma$. If $A$ is $M_2(K)$ then $\sigma( \left( \begin{smallmatrix} w & x\\ y & z \end{smallmatrix} \right) ) =\left( \begin{smallmatrix}z & {-x}\\ {-y} & w \end{smallmatrix} \right) $

@@

\TODO{Theorem expessing the $E$ splits $M$ condition}

@@

\begin{theorem}
Let $(A,\tau_A)$ be a central simple algebra of degree $n$ over $(E,\sigma)$ with involution of the second kind.
Suppose $(C,\tau_C)$ is a central simple algebra of degree $n$ over $k$ with orthogonal involution.
Then $(A,\tau_A)\hookrightarrow (C,\tau_C)$ if and only if there exists $(A_0,\tau_0)$ a central simple algebra over $k$ with syplectic involution and $(B,\tau_B)$ a quaternion algebra over $k$ with an orthogonal involution such that:
\begin{enumerate}
\item $(E,\sigma) \hookrightarrow (B,\tau_B)$.
\item $(A,\tau_A) = (A_0,\tau_0) \otimes_k (E,\sigma)$.
\item $(C,\tau_C) = (A_0,\tau_0) \otimes_k (B,\tau_B)$.
\end{enumerate}
\end{theorem}
\begin{proof}
\TODO{fill in sketch}

\TODO{Note-I suspect one can turn the argument about rataional conjugacy=rational isomorphism into an alternate constructive proof}
\begin{itemize}
\item
We know that $(A,\tau) = (A_0,\tau_0)\otimes (E,\sigma)$ for a unique $(A_0,\tau_0)$.
\item
Composition of maps then gives us an inclusion $(A_0,\tau_0) \hookrightarrow (A,\tau_A)$.
\item
It is then automatic that there exists $B$ such that $C= A_0 \otimes_k B$.
\item
Note that if $C$ split $A_0,B$ not split and $E$ split then no map $A_0\times A_0$ into $C$
 ($GL_{A_0\times A_0}$ has no representation of dimension $2n$, smallest is dimension $2n^2$
 this implies $E\hookrightarrow B$.
\item
We then can consider the restriction of $\tau_C$ to $A_0$ and $B$.
This determines $\tau_B'$ and $\tau_0'$.
\item
One of $\tau_B', \tau_0'$ is syplectic the other orthogonal.
We claim we can't get them backwords because the discriminant of $\tau_C$ would be $D(\tau_0')^2$ if we did.
In which case it is a square, and we can easily then swap $\tau_B'$ and $\tau_0'$ roles...
\end{itemize}
\end{proof}

\begin{proposition}
\TODO{Rational conjugacy classes}
Rational conjugacy is the same as rational isomorphism.
\end{proposition}
\begin{proof}
\TODO{this might also be modifiable to show that the above construction is universal}
\begin{itemize}
\item First claim any conjugate embedding (of groups) arises from a conjugate embedding (of algebras with involutions).
\item We want to show that any two embeddings $A \hookrightarrow C$ are conjugate.
        With $A,C$ as defined above but the embedding not necissarily coming from the condition.

@@

\TODO{Want to prove the following facts [that one hopes are facts]}
\begin{itemize}
\item The normalizer in $\Sp$ (resp. $\SO$) of $\U$ is $\mu_2 \rtimes \U$ (resp. $\U$).  [or something like this]
\subitem  suffices to check over the algebraic closure.
\item The centralizer of the center of $\U$ in $\Sp$ and $\SO$ is $\U$
\subitem suffices to check over the algebraic closure.
\subitem Note that the connected component of the normalizer of the center of $\U$ is the centralizer of the center.
              (first claim reduces to second + claim that normalizer is connected in $\SO$ case)
\end{itemize}
\end{proof}

\TODO{Invariants of orthogonal from those of unitary}

\begin{proposition}
Let $(A,\tau_A)$ be a central simple algebra of degree $n$ over $(E,\sigma)$ with involution of the second kind.
Suppose $(C,\tau_C)$ is a central simple algebra of degree $n$ over $k$ with orthogonal involution such that $(A,\tau_A)\hookrightarrow (C,\tau_C)$.

The invariants of $(A,\tau_A)$ and $(C,\tau_C)$ are related as follows:
\begin{enumerate}
\item The discriminant of $\tau_C$ is $\delta_E^{n}$.
\item The discriminant of the unitary involution should determine clifford of the orthogonal.
\TODO{phrasing of above... }
          changing discriminant (action of $d\in k^\times/N_{E/k}(E^\times)$ is (something like) by $(d,\delta_E)$ on class of clifford algebra.
\item The discriminant of the unitary involution should be determined (somehow) by discriminant of $A_0$
\TODO{Not sure if we want to say this in this proposition or a elsewhere}
\item The algebra $B$ is determined by $A_0$ and $C$. The discriminant of $\tau_B$ should be $\delta_E$.

\item The index of $\tau_A$ and $\tau_C$ are the same (or at least determine each other).
\end{enumerate}
\end{proposition}
\begin{proof}
\TODO{complete sketch}
\begin{itemize}
\item The first and fourth points follow from definition of discriminant (it is the norm of the appropriate element of $E$)
\item The third point is clear by construction but the actual dependence is important.
\item The second then follows from the third (by construction) but again, the actual dependence should be spelled out.
\item The fifth point in one direction is clear (maximal split in torus can be pushed forward). In the other direction is clear for $A,C$ matrix algebras. 
         Case by case analysis should cover the rest.
\end{itemize}
\end{proof}

Based on the above we conclude:
\begin{proposition}
\TODO{Summary - Given orthogonal which unitary are possible}
\begin{itemize}
\item The index is determined.
\item The algebra $E$ is determined when $n$ is odd.
\item When $n$ is even, the discriminant must be trivial or there are NO unitary groups.
\item The choice of $B$ determines $A_0$ which determines $A$.
\item The Clifford algebra determines the discriminant of the unitary group.
\end{itemize}
\end{proposition}

Based on the above we also conclude:
\begin{proposition}
\TODO{Summary - Given unitary which orthogonal are possible}
\begin{itemize}
\item The discriminant and index are determined. (non-trivial discriminants can never occur if $n$ is even).
\item The Clifford algebra can vary, this depends on the choice of $B$ in the construction.
\subitem The choice of $B$ determines $C$... the choice is then not really of the clifford algebra (as of $C$), which then together with the choice of discriminant of the unitary group determines the Clifford algebra.
\subitem The interpretation of the Clifford algebra as a cohomology class in some sense depends on choice of map $E\hookrightarrow C$, which depends on $E\hookrightarrow B$.
               The apparent difficulties of this are likely allieviated by the fact that $n$ even forces trivial discriminant!.

\item The limitations on choise are that $E$ splits $B$, this limits choices for $C$.
\end{itemize}
\end{proposition}

\TODO{same as above but for symplectic}

@@

\begin{itemize}
\item I think $T_{E,\sigma}$ which embeds in an orthogonal is in the image of a unitary if and only if $E$ has a quadratic subextension which $\sigma$ restricts non-trivially to.

Fixing $x$ a generator of this subextension, the unitary group should come from the centralizer of this in the orthogonal group (suffices to check over algebraic closure).

\item \TODO{example of an even degree field with no degree $2$ subextension - an alternating group Galois extension should work}

\item Any torus in the image from a unitary group should arise from a unique rational conjugacy class of unitary group (determined by the quadratic subextension).

\item I think the same is true for symplectic groups.

\item Given $T_{E,\sigma}$ not sure when it embeds in both a symplectic and orthgonal (for the same algebra).
\subitem
             It will if it embeds into a unitary (as it is then in the intersection).
\subitem
             But I suspect that every torus embedding in a symptlectic (resp orthogonal) embeds in an orthogonal (resp symplectic) though likely not into all the orthogonals.
\end{itemize}

\TODO{cohomological reinterpretation is now or later?}
   - This is largely the difference between the "weil groups" of the AN and DN/CN types.
          (really the weyl semi direct the outer automorphism group)
   -
